Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the NSF (details).}

### 2022-04-07 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
folder for ASL submission

\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{iris}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{bst}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}

@@

\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking}
\author{Roshan Sharma}{University of Illinois at Chicago, USA}{}{}{}
\author{Shengyi Wang}{Princeton University, USA}{}{}{}
\author{Alexander Oey}{, USA}{}{}{}
\author{Anastasiia Evdokimova}{University of Illinois at Chicago, USA}{}{}{}
\author{Lennart Beringer}{Princeton University, USA}{}{}{}
\author{William Mansky}{University of Illinois at Chicago, USA}{}{}{}

\authorrunning{R. Sharma, S. Wang, A. Oey, A. Evdokimova, L. Beringer, and W. Mansky} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Roshan Sharma, Shengyi Wang, Alexander Oey, Anastasiia Evdokimova, Lennart Beringer, and William Mansky} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{I want to thank \dots}%optional

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
abstract
\end{abstract}

@@

This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
\begin{align*}
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
&\texttt{acquire}(\ell');\\
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
&\texttt{release}(\ell);\\
&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{align*}
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
\begin{align*}
&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
&\texttt{acquire}(\ell');\\
&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
&\texttt{release}(\ell);\\
&\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{align*}
Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple

@@

\section{Locks and Atomicity}

@@

\subsubsection{Key Ranges}
When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree. To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree. As observed by Krishna et al.~\cite{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node. We refer to the product of these bounds as a \emph{range}. The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$. Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes. 
\begin{figure}[htb]
\centering
\includegraphics[width=70mm,scale=0.5]{../FIG/range_prop.png}
\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
\label{range_bst}
\end{figure}
At the leaves, these ranges partition the space of all possible keys. If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.  For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}. The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.

@@

The abstract state of a tree is a map from keys to values. The $\treerep$ predicate then describes the conditions under which a map $m$ is implemented by the ghost state of a tree. Roughly speaking, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$. We formalize this by first defining the collection of $\mathsf{public\_half}$ assertions---i.e., the authoritative copies of the per-node ghost states---needed to form an abstract tree:
\begin{align*}
\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
& |\ \mathsf{Leaf} \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{None})\\
& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{Some}\ (k, v, g_l, g_r)) * \mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
\end{align*}
We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{public\_half}$ assertions of each node into a single assertion. We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.

This collection of $\mathsf{public\_half}$ assertions can be combined with per-node lock variants to ensure that the nodes in an implementation actually form a valid tree. However, we also need to record the fact that each node in the implementation actually belongs to the tree $t$. We do this by augmenting $\mathsf{public\_halves}$ with a piece of ghost state $\mathsf{ghost\_nodes}(S)$ that records the set of all nodes in the tree, constructed so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$. Then our final definition of $\treerep$ is:
$$\treerep(m, g) \triangleq \exists t.\ t \text{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$. Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node names and key-value pairs. (We will take advantage of this property when verifying the \lstinline{delete} function.)

>>>>

\documentclass[runningheads]{llncs}

\bibliographystyle{splncs04}% the mandatory bibstyle

\usepackage{txfonts}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{cases}
\usepackage{hyperref}
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
\usepackage{color}
\renewcommand\UrlFont{\color{blue}\rmfamily}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{amssymb}
\usepackage{iris-llncs}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\us}{$\mu$s}
\newcommand{\gnamety}{\ensuremath{\mathsf{gname}}}
\newcommand{\treerep}{\ensuremath{\mathsf{bst\_abs}}}
\newcommand{\nodeboxrep}{\ensuremath{\mathsf{bst\_ref}}}

@@

\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking\thanks{Supported by the NSF (details).}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Roshan Sharma\inst{1} \and
Shengyi Wang\inst{2}\orcidID{?} \and
Alexander Oey\inst{3} \and
Anastasiia Evdokimova\inst{4} \and
Lennart Beringer\inst{2}\orcidID{?} \and
William Mansky\inst{4}\orcidID{0000-0002-5351-895X}}

\authorrunning{R. Sharma et al.}

\institute{Amazon, USA \and Princeton University, USA \and Cisco, USA \and University of Illinois Chicago, USA} %do we need a note about prior affiliations?

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} https://dl.acm.org/ccs/ccs_flat.cfm

@@

@@

\begin{abstract}
Modern concurrent separation logic techniques have been very effective at proving that lock-free data structures implement atomic specifications: they perform the same operations as sequential data structures, and do not expose any intermediate states to other threads. In this paper, we describe a method of proving the same kinds of specifications for lock-based data structures, with both coarse-grained and fine-grained locking patterns. We begin from the traditional separation logic specifications of locks and lock invariants, and combine them with a ghost state pattern that connects the critical section of a lock with the linearization point of a data structure operation. For fine-grained data structures, this allows us to decompose the proofs into a proof that a lock-based program performs some local changes to ghost state, and a proof that these local changes correctly implement an update to the global state of the data structure. Our proof technique is implemented in the Verified Software Toolchain and applied to C implementations of lock-based concurrent data structures.

\keywords{concurrent separation logic, fine-grained locking, logical atomicity, Verified Software Toolchain}
\end{abstract}

@@

This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
$$\begin{array}{l}
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{array}$$
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
$$\begin{array}{l}
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
\texttt{acquire}(\ell');\\
\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
\texttt{release}(\ell);\\
\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{array}$$
Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

@@

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple
%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t. the situation where one of the other threads' action following the linearization point removes the element just inserted.

@@

\section{Locks and Atomicity}
Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
\begin{mathpar}
\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathsf{insert}\ t\ k\ v)\}

\{\mathsf{BST}\ p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\ p\ t \land \mathsf{lookup}\ t\ k = v\}

\{\mathsf{BST}\ p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\ p\ (\mathsf{delete}\ t\ k)\}
\end{mathpar}
where $\mathsf{BST}\ p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$. For instance, $t$ may be a recursively defined tree type in Coq, and $\mathsf{insert}$ may be a pure Coq function that adds a key-value pair to such a tree; $\mathsf{BST}$ could then be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, and $p$ could be the location of the root node.

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs:
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: in separation logic terms, the $\mathsf{BST}$ assertion included full ownership of every location in the data structure. The easiest way to fix this is to add a lock that each operation acquires at the start and releases at the end, with $\mathsf{BST}\ p\ t$ as its \emph{lock invariant}: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.

In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock. Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$. This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure. For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$.

@@

\subsubsection{Key Ranges}
When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree. To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree. As observed by Krishna et al.~\cite{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node. We refer to the product of these bounds as a \emph{range}. The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$. Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes.

\begin{figure}[htb]
\centering
\includegraphics[width=70mm,scale=0.5]{../FIG/range_prop.png}
\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
\label{range_bst}
\end{figure}

\usetikzlibrary{positioning}
\forestset{
   lab/.style={
        label={[anchor=west, scale=0.75] north east:#1},
    },
}
\begin{forest}
for tree={
    grow=south,
    circle, draw, minimum size=3ex, inner sep=1pt,
    s sep=3mm
        }
    % my label/.style={
        % label={[anchor=south west] north east:#1},
    % },
    [ 20, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
        [15, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,20)$};}]
        [30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(20,\infty)$};}
            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(20,30)$};}]
            [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
                [45, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,45)$};}
                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}
                            [, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
                            [, tikz={\node[right=0pt of .north east, scale=0.75]  {$(35,40)$};}]
                        ]
                        [,no edge, draw=none]
                    ]
                    [,no edge, draw=none]
                ]
                [,no edge, draw=none]
            ]
        ]
    ]
\end{forest}

At the leaves, these ranges partition the space of all possible keys. If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.  For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}. The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.

@@

The abstract state of a tree is a map from keys to values. The $\treerep$ predicate then describes the conditions under which a map $m$ is implemented by the ghost state of a tree. Roughly speaking, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$. We formalize this by first defining the collection of $\mathsf{public\_half}$ assertions---i.e., the authoritative copies of the per-node ghost states---needed to form an abstract tree:
$$\begin{array}{r l}
\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
& |\ \mathsf{Leaf} \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{None})\\
& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{Some}\ (k, v, g_l, g_r)) * \mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
\end{array}$$
We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{public\_half}$ assertions of each node into a single assertion. We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.

This collection of $\mathsf{public\_half}$ assertions can be combined with per-node lock variants to ensure that the nodes in an implementation actually form a valid tree. However, we also need to record the fact that each node in the implementation actually belongs to the tree $t$. We do this by augmenting $\mathsf{public\_halves}$ with a piece of ghost state $\mathsf{ghost\_nodes}(S)$ that records the set of all nodes in the tree, constructed so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$. Then our final definition of $\treerep$ is:
$$\treerep(m, g) \triangleq \exists t.\ t \textrm{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$. Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node names and key-value pairs. (We will take advantage of this property when verifying the \lstinline{delete} function.)

As described in section~\ref{atomicity}, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock variant assertion $\ell \islock_\pi R$, where $R$ is parameterized by the abstract state of the node (range + possible contents). We will fix the share $\pi$ to 1/2 so that the range in a node's lock can be out of date with respect to its $\mathsf{public\_half}$ (in particular, it may have grown due to rearrangements further up the tree). We then need to choose a variant $R$ describing the resources held by the lock for a given abstract state. The C implementation of a node is:
\begin{lstlisting}[language = C,numbers = none]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
\end{lstlisting}
The node data of each \lstinline{tree} is wrapped in a \lstinline{tree_t} struct, and is only accessible when the lock in the \lstinline{lock} field is held. The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
$$\begin{array}{l}
\mathsf{match}\ c\ \mathsf{with}\\
|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
\end{array}$$
We then combine this with the hand-over-hand invariant pattern of section~\ref{selflock}, giving us a final lock variant of:
$$\begin{array}{r l}
\mathsf{node\_inv}((l, r), c) \triangleq \ &p.\texttt{lock} \lockvar_{\pi_1} \mathsf{node\_inv} * \mathsf{match}\ c\ \mathsf{with}\\
&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * \\&\qquad p_l.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv} * p_r.\texttt{lock} \lockvar_{\pi_2} \mathsf{node\_inv}
\end{array}$$
While we do not know the ranges or contents of the child nodes, the lock-variant construction guarantees that they are consistent with the $\mathsf{public\_half}$ assertions of the global ghost state, and the hand-over-hand pattern allows us to acquire $p_l.\texttt{lock}$ or $p_r.\texttt{lock}$ and then release $p.\texttt{lock}$ without losing any shares of the lock assertions.
